---
title: "Exercises — Linear algebra"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
    html-math-method: katex
    embed-resources: true
---

## Solving linear systems

- Wikipedia also summarises [the method of Gaussian elimination](https://en.wikipedia.org/wiki/Gaussian_elimination).

### Q1.1
Solve for $x$ and $y$ in: $x+y=1;\ x+2y=2$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Augmented matrix:
$$\left[\begin{array}{cc|c} 1 & 1 & 1 \\ 1 & 2 & 2 \end{array}\right]$$

1. $R_2 \leftarrow R_2 - R_1$:
$$\left[\begin{array}{cc|c} 1 & 1 & 1 \\ 0 & 1 & 1 \end{array}\right]$$
2. From row 2: $y = 1$.
3. Back-substitute: $x = 1 - y = 0$.

**Answer:** $x = 0,\ y = 1$.
:::

### Q1.2
Solve for $x$ and $y$ in: $x = 1;\ 4x + 4y = 0$ &nbsp;&nbsp; (note: you can still arrange the equations for Gaussian elimination, just leave a "hole" if a variable is missing).

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Augmented matrix (note the missing $y$ in eq. 1 → 0):
$$\left[\begin{array}{cc|c} 1 & 0 & 1 \\ 4 & 4 & 0 \end{array}\right]$$

1. $R_2 \leftarrow R_2 - 4 R_1$:
$$\left[\begin{array}{cc|c} 1 & 0 & 1 \\ 0 & 4 & -4 \end{array}\right]$$
2. From row 2: $y = -1$.
3. From row 1: $x = 1$.

**Answer:** $x = 1,\ y = -1$.
:::

### Q1.3
Solve for $x$ and $y$ in: $2x + 4y = 2;\ x + y = 0$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Augmented matrix:
$$\left[\begin{array}{cc|c} 2 & 4 & 2 \\ 1 & 1 & 0 \end{array}\right]$$

1. Swap rows for a friendlier pivot:
$$\left[\begin{array}{cc|c} 1 & 1 & 0 \\ 2 & 4 & 2 \end{array}\right]$$
2. $R_2 \leftarrow R_2 - 2 R_1$:
$$\left[\begin{array}{cc|c} 1 & 1 & 0 \\ 0 & 2 & 2 \end{array}\right]$$
3. $y = 1$, then $x = -y = -1$.

**Answer:** $x = -1,\ y = 1$.
:::

### Q1.4
Solve for $x$, $y$ and $z$ in: $3x + 3y + 3z = 3;\ x + 2y + 2z = 2;\ 2x + x + 2z = 2$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
The third equation simplifies to $3x + 2z = 2$. Augmented matrix:
$$\left[\begin{array}{ccc|c} 3 & 3 & 3 & 3 \\ 1 & 2 & 2 & 2 \\ 3 & 0 & 2 & 2 \end{array}\right]$$

1. $R_1 \leftarrow R_1 / 3$:
$$\left[\begin{array}{ccc|c} 1 & 1 & 1 & 1 \\ 1 & 2 & 2 & 2 \\ 3 & 0 & 2 & 2 \end{array}\right]$$
2. $R_2 \leftarrow R_2 - R_1$, $R_3 \leftarrow R_3 - 3 R_1$:
$$\left[\begin{array}{ccc|c} 1 & 1 & 1 & 1 \\ 0 & 1 & 1 & 1 \\ 0 & -3 & -1 & -1 \end{array}\right]$$
3. $R_3 \leftarrow R_3 + 3 R_2$:
$$\left[\begin{array}{ccc|c} 1 & 1 & 1 & 1 \\ 0 & 1 & 1 & 1 \\ 0 & 0 & 2 & 2 \end{array}\right]$$
4. Back-substitute: $z = 1$, $y = 1 - z = 0$, $x = 1 - y - z = 0$.

**Answer:** $x = 0,\ y = 0,\ z = 1$.
:::

### Q1.5
Solve for $x$, $y$ and $z$ in: $3x = 3;\ x + 2y + 2z = 0;\ 2x + 2z = 0$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
1. From eq. 1: $x = 1$.
2. Substitute into eq. 3: $2 + 2z = 0 \Rightarrow z = -1$.
3. Substitute into eq. 2: $1 + 2y - 2 = 0 \Rightarrow y = \tfrac{1}{2}$.

**Answer:** $x = 1,\ y = \tfrac{1}{2},\ z = -1$.
:::

If you want even more, see eq. 1–6 from <https://online.math.uh.edu/Math1310-unpaid/ch6/s61/ex61.pdf> but solve them as shown before. Don't worry about graphing etc.

### Q1.6
A nucleic-acid prep typically contains both DNA and protein contamination. Both absorb UV light, with different "extinction coefficients" $\varepsilon$ at 260 nm and 280 nm. Beer–Lambert says the absorbance at each wavelength is the sum of the contributions:
$$A_\lambda = \varepsilon^{\text{DNA}}_\lambda\,[\text{DNA}] + \varepsilon^{\text{prot}}_\lambda\,[\text{prot}]$$

Suppose (in convenient units) the extinction coefficients are

| species | 260 nm | 280 nm |
|---------|:------:|:------:|
| DNA     | 2      | 1      |
| Protein | 1      | 2      |

You measure $A_{260} = 5$ and $A_{280} = 4$. What are $[\text{DNA}]$ and $[\text{prot}]$?

> *Hint:* this gives two equations in two unknowns — solve by Gaussian elimination, exactly as in Q1.1–Q1.5.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Plug the measurements into Beer–Lambert at each wavelength:
$$\begin{aligned}
2\,[\text{DNA}] + 1\,[\text{prot}] &= 5 \\
1\,[\text{DNA}] + 2\,[\text{prot}] &= 4
\end{aligned}$$

Subtract twice eq. 2 from eq. 1: $-3\,[\text{prot}] = -3 \Rightarrow [\text{prot}] = 1$, then $[\text{DNA}] = 2$.

**Answer:** $[\text{DNA}] = 2$, $[\text{prot}] = 1$. The sample is mostly DNA with measurable protein contamination — essentially what a NanoDrop computes when it reports the $A_{260}/A_{280}$ ratio.
:::

## Practice using matrices and vectors

Watch the lecture video to see how these let you do the same math, but with a more compact notation. Wikipedia also summarises [matrices](https://en.wikipedia.org/wiki/Matrix_(mathematics)) and [vectors](https://en.wikipedia.org/wiki/Vector_space).

Write the following equation system using matrices and vectors:

### Q2.1
$5x + y = 1;\ x + 3y = 5$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Read coefficients of $x$ and $y$ as the rows of the matrix; the right-hand sides become the entries of $\mathbf{b}$:
$$\begin{bmatrix} 5 & 1 \\ 1 & 3 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} 1 \\ 5 \end{bmatrix}$$
:::

### Q2.2
$x + y + z = 5;\ 2x - z = 4;\ 3y + z = 2$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
A "missing" variable in an equation gets a 0 entry:
$$\begin{bmatrix} 1 & 1 & 1 \\ 2 & 0 & -1 \\ 0 & 3 & 1 \end{bmatrix} \begin{bmatrix} x \\ y \\ z \end{bmatrix} = \begin{bmatrix} 5 \\ 4 \\ 2 \end{bmatrix}$$
:::

Write this as a system of equations without using matrix notation:

### Q2.3
$$\begin{bmatrix} 1 & 2 \\ 4 & 8 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Each row of the matrix · the column vector gives one equation:
$$\begin{aligned} x + 2y &= 0 \\ 4x + 8y &= 1 \end{aligned}$$

Note the second equation is 4 times the first on the left, but not on the right — so this system is **inconsistent** (no solution). We will revisit this in Q3.
:::

### Q2.4
$$\begin{bmatrix} 1 & 0 & 1 \\ 2 & 2 & 0 \\ 3 & 2 & 0 \end{bmatrix} \begin{bmatrix} x \\ y \\ z \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 2 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
$$\begin{aligned} x + z &= 0 \\ 2x + 2y &= 1 \\ 3x + 2y &= 2 \end{aligned}$$
:::

### Q2.5
$$\begin{bmatrix} 1 & 2 \end{bmatrix} \begin{bmatrix} x \\ x^2 \end{bmatrix} = \begin{bmatrix} 3 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
$$x + 2x^2 = 3$$

Note this is **not linear** in $x$ (because of $x^2$). Matrix notation with a non-linear "variable" inside the vector is a common trick, but Gaussian elimination won't work directly here.
:::

Perform the matrix multiplication:

### Q2.6
$$\begin{bmatrix} 3 & 4 \end{bmatrix} \begin{bmatrix} 2 \\ 5 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Inner product: $3 \cdot 2 + 4 \cdot 5 = 6 + 20 = 26$.
:::

Solve the following equations using the shorthand notation provided in the lecture:

### Q2.7
$$\begin{bmatrix} 4 & -5 \\ 3 & 4 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} 24 \\ -13 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Augmented matrix:
$$\left[\begin{array}{cc|c} 4 & -5 & 24 \\ 3 & 4 & -13 \end{array}\right]$$

1. $R_2 \leftarrow 4 R_2 - 3 R_1$ (clears $x$ from row 2):
$$\left[\begin{array}{cc|c} 4 & -5 & 24 \\ 0 & 31 & -124 \end{array}\right]$$
2. $y = -124/31 = -4$.
3. $4x - 5(-4) = 24 \Rightarrow 4x = 4 \Rightarrow x = 1$.

**Answer:** $x = 1,\ y = -4$.
:::

### Q2.8
Verify the solutions to the equation systems so far using [the R `solve` function](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/solve).

::: {.callout-tip collapse="true" title="Step-by-step solution"}
```r
# Q1.1
A <- matrix(c(1, 1,
              1, 2), 2, 2, byrow = TRUE)
solve(A, c(1, 2))           # 0, 1

# Q1.4
A <- matrix(c(3, 3, 3,
              1, 2, 2,
              3, 0, 2), 3, 3, byrow = TRUE)
solve(A, c(3, 2, 2))        # 0, 0, 1

# Q2.7
A <- matrix(c(4, -5,
              3,  4), 2, 2, byrow = TRUE)
solve(A, c(24, -13))        # 1, -4
```
Tip: `byrow = TRUE` is important — by default R fills matrices column by column.
:::

### Q2.9
Rewrite the DNA-vs-protein problem from Q1.6 in matrix form $E\,\mathbf{c} = \mathbf{A}$, and solve it in R.

> *Hint:* this is exactly the kind of system Q2.8 asked you to solve with R's `solve` function.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
The rows of $E$ are wavelengths (260, 280 nm) and the columns are species (DNA, protein):
$$\underbrace{\begin{bmatrix} 2 & 1 \\ 1 & 2 \end{bmatrix}}_{E}\,\underbrace{\begin{bmatrix} [\text{DNA}] \\ [\text{prot}] \end{bmatrix}}_{\mathbf{c}} = \underbrace{\begin{bmatrix} 5 \\ 4 \end{bmatrix}}_{\mathbf{A}}$$
```r
E <- matrix(c(2, 1,
              1, 2), 2, 2, byrow = TRUE)
solve(E, c(5, 4))     # 2, 1
```
For three species (e.g. add RNA), measure at a third wavelength and solve a 3×3 system the same way — the matrix framing scales with no extra effort, while the row-by-row Gaussian elimination from Q1.6 gets tedious fast.
:::

## Existence of unique solutions and rank

- Wikipedia [summary of rank](https://en.wikipedia.org/wiki/Rank_(linear_algebra)).
- You can compute rank in R using [`rankMatrix`](https://www.rdocumentation.org/packages/Matrix/versions/1.6-4/topics/rankMatrix). It is in the `Matrix` library — run `library(Matrix)` first.

**Quick rule of thumb (for an $n \times n$ system $A\mathbf{x} = \mathbf{b}$):**

- If $\operatorname{rank}(A) = \operatorname{rank}([A|\mathbf{b}]) = n$ → unique solution.
- If $\operatorname{rank}(A) = \operatorname{rank}([A|\mathbf{b}]) < n$ → infinitely many solutions.
- If $\operatorname{rank}(A) < \operatorname{rank}([A|\mathbf{b}])$ → no solution (inconsistent).

For each of Q3.1–Q3.4: determine whether the equation can be solved or not; if it can, provide the solution.

### Q3.1
$$\begin{bmatrix} 2 & -5 \\ 6 & -15 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} 10 \\ 13 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Row 2 of $A$ is exactly $3 \times$ row 1, so $\operatorname{rank}(A) = 1$.
But the right-hand side: $3 \cdot 10 = 30 \ne 13$. So in the augmented matrix, $\operatorname{rank}([A|\mathbf{b}]) = 2$.

Since $1 < 2$ → **no solution** (inconsistent).
:::

### Q3.2
$$\begin{bmatrix} 3 & 5 \\ 11 & 7 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix} = \begin{bmatrix} 11 \\ 7 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
$\det = 3\cdot 7 - 5\cdot 11 = 21 - 55 = -34 \ne 0$, so $\operatorname{rank}(A) = 2 = n$ → **unique solution**.

Solve it:

1. $11 R_1 - 3 R_2$: $11(3x+5y) - 3(11x+7y) = 11\cdot 11 - 3\cdot 7$ → $34y = 100$ → $y = 50/17$.
2. $3x + 5\cdot 50/17 = 11$ → $3x = (187 - 250)/17 = -63/17$ → $x = -21/17$.

**Answer:** $x = -21/17,\ y = 50/17$.
:::

### Q3.3
$$\begin{bmatrix} 1 & 9 & 3 \\ 3 & 3 & 3 \\ 1 & 3 & 3 \end{bmatrix} \begin{bmatrix} x \\ y \\ z \end{bmatrix} = \begin{bmatrix} 1 \\ 0 \\ 0 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Compute $\det$ along the top row:
$$\det = 1(3\cdot3 - 3\cdot3) - 9(3\cdot3 - 3\cdot1) + 3(3\cdot3 - 3\cdot1) = 0 - 54 + 18 = -36 \ne 0$$
→ **unique solution**.

Solving (eq. 2 − eq. 3 gives $2x = 0 \Rightarrow x = 0$, then eq. 2 gives $y + z = 0$, and eq. 1 gives $9y + 3z = 1 \Rightarrow 6y = 1$):

**Answer:** $x = 0,\ y = 1/6,\ z = -1/6$.
:::

### Q3.4
$$\begin{bmatrix} 1 & 0 & 0 \\ 1 & 1 & 1 \\ 2 & 1 & 1 \end{bmatrix} \begin{bmatrix} x \\ y \\ z \end{bmatrix} = \begin{bmatrix} 5 \\ 0 \\ 0 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Note row 3 = row 1 + row 2 in $A$, so $\operatorname{rank}(A) = 2 < 3$.

Check the right-hand side: row 1 + row 2 of $\mathbf{b}$ = $5 + 0 = 5$, but row 3 of $\mathbf{b}$ = $0$. Inconsistent! So $\operatorname{rank}([A|\mathbf{b}]) = 3$.

Since $2 < 3$ → **no solution**.
:::

### Q3.5
Verify your answers by computing the rank in R.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
```r
library(Matrix)

# Q3.1
A <- matrix(c(2,-5, 6,-15), 2, 2, byrow = TRUE)
b <- c(10, 13)
rankMatrix(A)[1]              # 1
rankMatrix(cbind(A, b))[1]    # 2  → no solution

# Q3.4
A <- matrix(c(1,0,0, 1,1,1, 2,1,1), 3, 3, byrow = TRUE)
b <- c(5, 0, 0)
rankMatrix(A)[1]              # 2
rankMatrix(cbind(A, b))[1]    # 3  → no solution
```
:::

### Q3.6
What happens if you try to use R `solve()` on an unsolvable system?

::: {.callout-tip collapse="true" title="Step-by-step solution"}
```r
A <- matrix(c(2,-5, 6,-15), 2, 2, byrow = TRUE)
solve(A, c(10, 13))
# Error in solve.default(A, c(10, 13)) :
#   Lapack routine dgesv: system is computationally singular
```
R throws an error because the matrix is singular (zero determinant / rank-deficient). It does **not** silently return a wrong answer — but it also does not distinguish "infinitely many solutions" from "no solution"; you have to check ranks for that.
:::

**Optional further exercises:** <https://online.math.uh.edu/Math1310-unpaid/ch6/s61/ex61.pdf>

### Q3.7
Two scenarios in which the spectrophotometry setup of Q1.6/Q2.9 fails:

a. You only record a *single* absorbance, $A_{260}$, but still want both $[\text{DNA}]$ and $[\text{prot}]$. Why is this impossible? What is the rank of the relevant matrix?
b. You measure at two wavelengths, but the two species you're trying to distinguish happen to have *proportional* extinction spectra (one species' row of $E$ is a scalar multiple of the other's). Why can't you separate them, no matter how careful you are?

> *Hint:* count rows of $E$ vs. unknowns, and check linear dependence — the rank-vs-unknowns logic from Q3.1–Q3.4.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
**a.** With one wavelength, $E$ is a $1 \times 2$ matrix and so $\operatorname{rank}(E) = 1 < 2$. One equation in two unknowns is underdetermined — infinitely many concentration pairs match any single measurement.

**b.** Two wavelengths give a $2 \times 2$ $E$, but if the rows are proportional they are linearly dependent and $\operatorname{rank}(E) = 1$ again. The matrix is singular: `solve(E, A)` will fail and there is no unique decomposition.

Rank deficiency in $E$ is the math telling you the experimental design can't answer the biological question — this is essentially why dsDNA and ssDNA can't be distinguished by UV absorbance alone (their spectra are too similar), and why labs reach for orthogonal methods (gel electrophoresis, fluorescent dyes) for that job.
:::

## Matrix multiplication

- Wikipedia: [matrix multiplication](https://en.wikipedia.org/wiki/Matrix_multiplication).

### Q4.1
Compute $AB$ and $BA$ for
$$A = \begin{bmatrix} 1 & 0 \\ 5 & 0 \end{bmatrix},\quad B = \begin{bmatrix} 3 & 2 \\ 4 & 3 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
Entry $(i,j)$ of the product = row $i$ of the left matrix · column $j$ of the right matrix.

$$AB = \begin{bmatrix} 1\cdot 3 + 0\cdot 4 & 1\cdot 2 + 0\cdot 3 \\ 5\cdot 3 + 0\cdot 4 & 5\cdot 2 + 0\cdot 3 \end{bmatrix} = \begin{bmatrix} 3 & 2 \\ 15 & 10 \end{bmatrix}$$

$$BA = \begin{bmatrix} 3\cdot 1 + 2\cdot 5 & 3\cdot 0 + 2\cdot 0 \\ 4\cdot 1 + 3\cdot 5 & 4\cdot 0 + 3\cdot 0 \end{bmatrix} = \begin{bmatrix} 13 & 0 \\ 19 & 0 \end{bmatrix}$$

So $AB \ne BA$ in general — matrix multiplication is **not commutative**.
:::

### Q4.2
Compute $AB$ and $BA$ for
$$A = \begin{bmatrix} 3 & 5 \\ 11 & 7 \end{bmatrix},\quad B = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}$$

Note that $B$ is the **identity matrix**, frequently written $\mathbb{I}$ or more precisely $\mathbb{I}_{2\times 2}$.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
$$AB = \begin{bmatrix} 3\cdot1 + 5\cdot 0 & 3\cdot 0 + 5\cdot 1 \\ 11\cdot 1 + 7\cdot 0 & 11\cdot 0 + 7\cdot 1 \end{bmatrix} = \begin{bmatrix} 3 & 5 \\ 11 & 7 \end{bmatrix} = A$$

$$BA = \begin{bmatrix} 1\cdot 3 + 0\cdot 11 & 1\cdot 5 + 0\cdot 7 \\ 0\cdot 3 + 1\cdot 11 & 0\cdot 5 + 1\cdot 7 \end{bmatrix} = \begin{bmatrix} 3 & 5 \\ 11 & 7 \end{bmatrix} = A$$

The identity matrix leaves any compatibly-sized matrix unchanged — like multiplying by 1.
:::

### Q4.3
Compute $AB$ for
$$A = \begin{bmatrix} 1 & 2 \\ 3 & 4 \end{bmatrix},\quad B = \begin{bmatrix} 5 & 5 \\ 4 & 4 \end{bmatrix}$$

::: {.callout-tip collapse="true" title="Step-by-step solution"}
$$AB = \begin{bmatrix} 1\cdot 5 + 2\cdot 4 & 1\cdot 5 + 2\cdot 4 \\ 3\cdot 5 + 4\cdot 4 & 3\cdot 5 + 4\cdot 4 \end{bmatrix} = \begin{bmatrix} 13 & 13 \\ 31 & 31 \end{bmatrix}$$

The two columns of $B$ are equal, so the two columns of $AB$ are also equal.
:::

### Q4.4
Verify your answers using R. You can simply do `A %*% B` to perform a multiplication.

Note that this is not the same thing **in R** as `A * B`, which does the math element-by-element! (which is also useful sometimes). R (and Python) is not always consistent with common mathematical notation; mathematicians might instead write element-wise multiplication as $A \odot B$ or other non-standard notation. These sorts of confusions are common throughout programming, and something one just has to get used to.

::: {.callout-tip collapse="true" title="Step-by-step solution"}
```r
A <- matrix(c(1, 0, 5, 0), 2, 2, byrow = TRUE)
B <- matrix(c(3, 2, 4, 3), 2, 2, byrow = TRUE)

A %*% B    # Q4.1: matrix product
#       [,1] [,2]
# [1,]    3    2
# [2,]   15   10

B %*% A    # Q4.1 reversed
#       [,1] [,2]
# [1,]   13    0
# [2,]   19    0

A * B      # element-wise — totally different operation!
#       [,1] [,2]
# [1,]    3    0
# [2,]   20    0
```
:::

## Extra exercise material

A second pass at the same material, for extra practice.

### Matrix multiplication and dimensions (extra)

- **QE1.1** Without using R, compute $AB$ for
$$A = \begin{bmatrix} 2 & 1 \\ 0 & 3 \\ 1 & 4 \end{bmatrix},\quad B = \begin{bmatrix} 1 & 2 & 0 \\ 3 & 1 & 1 \end{bmatrix}$$
What are the dimensions of the result? Then check in R with `%*%`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
$A$ is $3 \times 2$ and $B$ is $2 \times 3$. The "inner" dimensions match (both 2), so the product is defined and has the "outer" dimensions — $3 \times 3$.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
A <- matrix(c(2, 1,
              0, 3,
              1, 4), 3, 2, byrow = TRUE)
B <- matrix(c(1, 2, 0,
              3, 1, 1), 2, 3, byrow = TRUE)
A %*% B
```
By hand, row 1 of $A$ times the columns of $B$ gives $(2\cdot 1 + 1\cdot 3,\ 2\cdot 2 + 1\cdot 1,\ 2\cdot 0 + 1\cdot 1) = (5, 5, 1)$, and similarly for the other rows.
:::

- **QE1.2** What happens if you try `B %*% A` instead, where $A$ and $B$ are as in QE1.1? What are the dimensions of the result? What if you try `A %*% A`?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
For $X \cdot Y$ to be defined, the number of columns of $X$ must equal the number of rows of $Y$.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
A <- matrix(c(2, 1,
              0, 3,
              1, 4), 3, 2, byrow = TRUE)
B <- matrix(c(1, 2, 0,
              3, 1, 1), 2, 3, byrow = TRUE)
# B is 2x3 and A is 3x2 — inner dims match (3) → result is 2x2
B %*% A
# A is 3x2, A is 3x2 — inner dims do NOT match (2 vs 3) → error
try(A %*% A)
```
The take-away: matrix multiplication is picky about dimensions. `A %*% A` only makes sense if $A$ is square.
:::

- **QE1.3** Build the $4 \times 4$ identity matrix in R using `diag()`. Then verify that for any compatibly-sized matrix $M$, $\mathbb{I}\,M = M\,\mathbb{I} = M$. Try it on
$$M = \begin{bmatrix} 1 & 2 & 3 & 4 \\ 5 & 6 & 7 & 8 \\ 9 & 10 & 11 & 12 \\ 13 & 14 & 15 & 16 \end{bmatrix}$$

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`diag(4)` returns the $4 \times 4$ identity matrix. `diag(c(1,2,3))` builds a diagonal matrix with the given entries.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
I4 <- diag(4)
I4
M  <- matrix(1:16, 4, 4, byrow = TRUE)
I4 %*% M
M  %*% I4
# Both should equal M
all(I4 %*% M == M)
all(M  %*% I4 == M)
```
:::

### Solving $A\mathbf{x} = \mathbf{b}$ (extra)

- **QE2.1** Solve in R:
$$\begin{bmatrix} 1 & 2 & 0 \\ 3 & -1 & 1 \\ 0 & 1 & 4 \end{bmatrix} \begin{bmatrix} x \\ y \\ z \end{bmatrix} = \begin{bmatrix} 5 \\ 2 \\ 9 \end{bmatrix}$$
Verify your solution by computing $A\mathbf{x}$ and comparing to $\mathbf{b}$.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Build `A` with `matrix(..., byrow = TRUE)`, then call `solve(A, b)`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
A <- matrix(c(1,  2, 0,
              3, -1, 1,
              0,  1, 4), 3, 3, byrow = TRUE)
b <- c(5, 2, 9)
x <- solve(A, b)
x
# Verify
A %*% x
```
:::

- **QE2.2** A small mixing problem. Three reagent stocks (S1, S2, S3) contain different concentrations of three components (a, b, c):
$$\text{stock matrix } S = \begin{bmatrix} 1 & 0 & 2 \\ 2 & 1 & 0 \\ 0 & 3 & 1 \end{bmatrix}$$
(rows are components a/b/c, columns are stocks S1/S2/S3). You want to prepare a mix containing $(a,b,c) = (4, 5, 7)$. How many volumes of each stock do you need? Solve with R.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Let $\mathbf{v} = (v_1, v_2, v_3)$ be the volumes of each stock. Then $S\,\mathbf{v} = (4, 5, 7)$.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
S <- matrix(c(1, 0, 2,
              2, 1, 0,
              0, 3, 1), 3, 3, byrow = TRUE)
b <- c(4, 5, 7)
v <- solve(S, b)
v
# Sanity check
S %*% v
```
The framing is the same as the DNA/protein problem (Q2.9) — just with three components instead of two.
:::

### Determinants and rank (extra)

- **QE3.1** Compute the determinant of each matrix below (by hand, then verify with R's `det()`). For which of them does $A\mathbf{x} = \mathbf{b}$ have a unique solution for every $\mathbf{b}$?
$$A_1 = \begin{bmatrix} 2 & 3 \\ 1 & 4 \end{bmatrix},\quad A_2 = \begin{bmatrix} 2 & 4 \\ 3 & 6 \end{bmatrix},\quad A_3 = \begin{bmatrix} 1 & 0 & 2 \\ 0 & 1 & 1 \\ 1 & 1 & 3 \end{bmatrix}$$

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
$A\mathbf{x} = \mathbf{b}$ has a unique solution iff $\det(A) \ne 0$ (equivalently, $A$ is invertible / full rank).
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
For a $2\times 2$ matrix, $\det = ad - bc$. For $A_3$, expand along the first row, or note that row 3 = row 1 + row 2.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
A1 <- matrix(c(2, 3,
               1, 4), 2, 2, byrow = TRUE)
A2 <- matrix(c(2, 4,
               3, 6), 2, 2, byrow = TRUE)
A3 <- matrix(c(1, 0, 2,
               0, 1, 1,
               1, 1, 3), 3, 3, byrow = TRUE)
det(A1)  # 2*4 - 3*1 = 5
det(A2)  # 2*6 - 4*3 = 0  (rows proportional)
det(A3)  # 0  (row 3 = row 1 + row 2)
```
Only $A_1$ has a non-zero determinant, so only $A_1\mathbf{x} = \mathbf{b}$ is guaranteed to have a unique solution for every $\mathbf{b}$. $A_2$ and $A_3$ are singular — they have either no solution or infinitely many, depending on $\mathbf{b}$.
:::

- **QE3.2** Use `Matrix::rankMatrix()` to confirm your answers in QE3.1. For the singular cases, construct one right-hand side $\mathbf{b}$ that yields no solution and one that yields infinitely many.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
For $A_2$: row 2 is $1.5\times$ row 1. So $\mathbf{b}$ is consistent only when $b_2 = 1.5\, b_1$. For example $\mathbf{b} = (2, 3)$ is consistent (infinitely many solutions), but $\mathbf{b} = (1, 3)$ is not.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
library(Matrix)
A2 <- matrix(c(2, 4,
               3, 6), 2, 2, byrow = TRUE)
rankMatrix(A2)[1]                        # 1
rankMatrix(cbind(A2, c(2, 3)))[1]        # 1 → infinitely many solutions
rankMatrix(cbind(A2, c(1, 3)))[1]        # 2 → no solution
# solve() will fail on either:
try(solve(A2, c(2, 3)))
```
The rank rule from Section 3 still applies: $\operatorname{rank}(A) = \operatorname{rank}([A|\mathbf{b}]) = n$ → unique; equal but less than $n$ → infinitely many; unequal → none.
:::

### Bioinformatics-flavoured (extra)

- **QE4.1** Spectrophotometry, three species. Extend the DNA/protein setup of Q2.9 to also include RNA. You measure absorbance at three wavelengths (260, 280, 230 nm). The extinction coefficient matrix is
$$E = \begin{bmatrix} 2 & 1 & 2 \\ 1 & 2 & 1 \\ 3 & 1 & 2 \end{bmatrix}$$
(rows = wavelengths, columns = DNA / protein / RNA). You measure $\mathbf{A} = (8, 6, 9)$. Find the concentrations using R. Then check what happens if your last row of $E$ were instead $(3, 3, 3)$ — does the problem still have a unique solution?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Same setup as Q2.9, just bigger: solve $E\,\mathbf{c} = \mathbf{A}$. For the "what if" part, check `det(E)` or `rankMatrix(E)`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
E <- matrix(c(2, 1, 2,
              1, 2, 1,
              3, 1, 2), 3, 3, byrow = TRUE)
A <- c(8, 6, 9)
solve(E, A)
det(E)

# Now the alternative — last row replaced by (3, 3, 3)
E2 <- matrix(c(2, 1, 2,
               1, 2, 1,
               3, 3, 3), 3, 3, byrow = TRUE)
det(E2)
library(Matrix)
rankMatrix(E2)[1]
```
With the original $E$, the system has a unique solution. Replacing the last row with $(3, 3, 3)$ doesn't *immediately* look problematic, but $\det(E_2) = 0$ — the three rows are linearly dependent (in fact the last row equals row 1 + row 2). The third wavelength gives no new information, and DNA/protein/RNA cannot be separated from these measurements. This is the multi-species analogue of Q3.7b: rank deficiency in the design matrix means the experiment can't answer the biological question.
:::

