---
title: "Working with variables"
format:
  html:
    theme: yeti
    css: ../../../module2/module2_exercises_R.css
---

##  🎯 Objective

Practice fundamental R skills by creating variables of different types, exploring variable types, performing type casting, and other basic operations.

> In all exercises during this course, it is important that you try to figure out what the expected result would be, prior to running the commands. You should then verify that this will indeed be the result by running the command. In case there is a discrepency between your expectations and the actual output make sure you understand why before you move forward. If you cannot figure out how to, or which command to run you can click the key to reveal example code including expected output.

Also note that in many cases there are multiple solutions that solve the problem equally well.

## 📝 Instructions

#### **Set up new project**
1. Open RStudio on your computer
2. Create a new project:
    - Click on "File" in the top menu
    - Select "New Project" and choose "New Directory."
    - Choose "New Project."
    - Enter a project name (e.g., "VariableManipulation") and specify the project directory
    - Click "Create Project."
    - Make sure you have the 4 panel view, with the code editor visible



#### **Verify working directory**

In R, `getwd()` stands for "get working directory," and `setwd()` stands for "set working directory." The working directory is the location on your computer where R looks for and saves files by default. When starting a new project, the working directory will be set to that folder by default, but it's still good practice to check the working directory at the beginning of your R script or project.

Make sure your working directory is in the folder you specified for the project:

```{r, eval=FALSE}
getwd()
```

Make it a habit to always write your code in the editor panel, and run it by highlighting it and press run. This way all your code will be saved, and an easily be re-run.

#### **Working with variables**

Assign values to the following variables:

```{r, echo=TRUE}
age <- 25
name <- "John"
is_student <- TRUE
```

Use the `typeof()` and `class()` functions to find out the types of the variables:

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=FALSE}
typeof(age)
class(age)
typeof(name)
class(name)
typeof(is_student)
class(is_student)
```
::: 

**Type Casting**  

Convert the age variable to a character type using the as.character() function.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
age_as_string <- as.character(age)
```
::: 

**Concatenate character variables**

Sometimes you would want to add together two variables of the type character, concatenating them. For this we can use the `paste()` and `paste0()` function:

Concatenate the `name` and `age_as_string` variables using both paste() and paste0() functions:

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
full_message_paste <- paste("My name is", name, "and I am", age_as_string, "years old.")
full_message_paste0 <- paste0("My name is", name, "and I am", age_as_string, "years old.")
```
::: 

**Print Formatted Text**

Use the `print()` function to display the `full_message_paste` and `full_message_paste0`.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
#| code-fold: false
print(full_message_paste)
print(full_message_paste0)
```
:::


What seems to be the difference between `paste()` and `paste0()`?

Find out more by reading the documentation for `paste()` and `paste0()`: 

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=FALSE}
?paste()
```
::: 

Try adding a separator as per the documentation, and see what happens.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
full_message_paste_sep <- paste("My name is", name, "and I am", age_as_string, "years old.", sep = ':')
print(full_message_paste_sep)
```
::: 

Inspect the Environment panel to see the listing of all the variables you have saved in memory so far. Try removing all environment variables by clicking on the small broom icon and see what happens.

## Extra exercise material

A second pass at the same material, for extra practice.

### Variables and types (extra)

- **QE1.1** Create three variables: `height_cm` with the value `172.5`, `gene_symbol` with the value `"TP53"`, and `is_expressed` with the value `FALSE`. For each variable, predict what `typeof()` and `class()` will report before running the commands.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Numbers with decimals are usually stored as `"double"` in R. Text in quotes becomes `"character"`. `TRUE`/`FALSE` becomes `"logical"`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
height_cm <- 172.5
gene_symbol <- "TP53"
is_expressed <- FALSE

typeof(height_cm)
class(height_cm)
typeof(gene_symbol)
class(gene_symbol)
typeof(is_expressed)
class(is_expressed)
```
:::

- **QE1.2** In R, a plain number like `5` is stored as a "double" by default, but you can force it to be an "integer" by appending `L` (e.g. `5L`). Create `n_samples <- 10` and `n_replicates <- 10L`. Inspect `typeof()` for both. Then create `treatment <- factor("control")` and compare `typeof(treatment)` to `class(treatment)`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`typeof()` returns the underlying storage mode, while `class()` returns the higher-level R class. For factors these differ.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
n_samples <- 10
n_replicates <- 10L
typeof(n_samples)
typeof(n_replicates)

treatment <- factor("control")
typeof(treatment)
class(treatment)
```
:::

### Type casting (extra)

- **QE2.1** Start with the character variable `count_string <- "42"`. Convert it to a numeric using `as.numeric()`, then to an integer using `as.integer()`. Verify the types using `typeof()` after each conversion.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
You can store the result in a new variable each step, e.g. `count_num <- as.numeric(count_string)`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
count_string <- "42"
count_num <- as.numeric(count_string)
count_int <- as.integer(count_string)

typeof(count_num)
typeof(count_int)
count_num
count_int
```
:::

- **QE2.2** What happens if you try to cast a non-numeric string like `"abc"` to a number? Try `as.numeric("abc")` and observe both the result and the message R prints.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
R will not throw an error, but it will produce a special value indicating "not available". Look in the console for a warning.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
bad_cast <- as.numeric("abc")
bad_cast
is.na(bad_cast)
```
The result is `NA`, and R prints a warning: `NAs introduced by coercion`. This is R's gentle way of telling you the cast failed.
:::

### String manipulation (extra)

- **QE3.1** Create two variables `first <- "BRCA"` and `second <- "1"`. Use `paste0()` to combine them into a single gene name (no space), and then use `paste()` with `sep = "-"` to combine them as `"BRCA-1"`. Print both results.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`paste0()` is equivalent to `paste()` with `sep = ""`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
first <- "BRCA"
second <- "1"

gene_name <- paste0(first, second)
gene_name_dashed <- paste(first, second, sep = "-")

print(gene_name)
print(gene_name_dashed)
```
:::

- **QE3.2** Take the string `dna <- "ACGTACGT"`. Use `nchar()` to find out how many characters it contains, `tolower()` to convert it to lowercase, and `toupper()` to convert it back to uppercase.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Each function takes the string as a single argument and returns a new value — it does not modify the original variable.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
dna <- "ACGTACGT"
nchar(dna)
tolower(dna)
toupper(tolower(dna))
```
:::

- **QE3.3** Given `seq <- "ATGCGTACGTTAG"`, use `substr()` to extract the first three characters (which would be the start codon). Then extract characters 4 to 6.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`substr(x, start, stop)` returns the substring from position `start` to position `stop`, inclusive. R uses 1-based indexing.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
seq <- "ATGCGTACGTTAG"
start_codon <- substr(seq, 1, 3)
next_codon <- substr(seq, 4, 6)

print(start_codon)
print(next_codon)
```
:::

### Bioinformatics-flavoured cast (extra)

- **QE4.1** In many bioinformatics datasets, binary outcomes (e.g. whether a sample is a case or control) are stored as `"1"` or `"0"` strings. Create the vector `outcome <- c("1", "0", "1", "1", "0")` and convert it to a logical vector where `"1"` becomes `TRUE` and `"0"` becomes `FALSE`. Verify the result with `typeof()`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
You can first cast the character vector to numeric using `as.numeric()`, and then cast that to logical using `as.logical()`. In R, `1` is treated as `TRUE` and `0` as `FALSE`.
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Try chaining the two casts: `as.logical(as.numeric(outcome))`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
outcome <- c("1", "0", "1", "1", "0")
outcome_logical <- as.logical(as.numeric(outcome))

print(outcome_logical)
typeof(outcome_logical)
```
:::
