---
title: "Control flows in R"
format:
  html:
    theme: yeti
    css: ../../../module2/module2_exercises_bash.css
---

##  🎯 Objective

Practice iterating over sequences using loops and execute different code depending on condition.

> You will not be tested on control flows for Module 2, BUT you will need these skills later for Module 4. So take some time to work through these exercises, in preparation for that module.

## 🧮 Additional reading (Optional)

There are many resources out there if you want to further dig into loops, if/else statements, and functions. Here is a list of a few *optional* resources for you to dig deeper into:

**for loops:**   

* [https://datacarpentry.org/semester-biology/materials/for-loops-R/](https://datacarpentry.org/semester-biology/materials/for-loops-R/) 
* [https://www.w3schools.com/r/r_for_loop.asp](https://www.w3schools.com/r/r_for_loop.asp)
* [https://www.youtube.com/watch?v=4-uWgh5kDSc](https://www.youtube.com/watch?v=4-uWgh5kDSc)

**while loops:**  

* [https://www.w3schools.com/r/r_while_loop.asp](https://www.w3schools.com/r/r_while_loop.asp)

**if/else statements:**  

* [https://www.w3schools.com/r/r_if_else.asp](https://www.w3schools.com/r/r_if_else.asp)

**Functions:**  

* [https://www.youtube.com/watch?v=3uK1OzA7CTs](https://www.youtube.com/watch?v=3uK1OzA7CTs)
* [https://www.w3schools.com/r/r_functions.asp](https://www.w3schools.com/r/r_functions.asp)
* [https://datacarpentry.org/semester-biology/materials/functions-R/](https://datacarpentry.org/semester-biology/materials/functions-R/)


## 📝 Instructions

### Loops

```{r}
my_vec <- c(2, 4 ,7, 12, 5, 23)
```

Use a `for` loop to iterate over the vector above. Start by, in the loop, just printing the number.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
for (val in my_vec) {
    print(val)
}
```
::: 

Next, count how many values are in the list by using a counter to count each iteration. Remember to initiate the counter outside the loop (check what happens if you initiate the counter inside the loop!).

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
counter <- 0
for (val in my_vec) {
    counter <- counter + 1
}
print(counter)
```
::: 

Now when we are warmed up, let's work on some more interesting data! Download [this file](https://www.canvas.umu.se/files/3925284/download?download_frd=1) and use `read.csv()` to read it into a dataframe. Don't forget to use the path where you have the file. Always start with doing a quick check how the data is structured.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=TRUE}
data <- read.csv("data/diabetes_cleaned.csv")
str(data)
```
::: 

Start by looping over the *rows* of the table, print the value of the first column. Remember, the default is to loop over columns, not rows.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=F}
for (i in 1:nrow(data)) {
    print(data[i,1])
}
```
```{r, eval=T, echo=F}
for (i in 1:nrow(data)) {
    print(data[i,1])
    if (i == 5) {
        break
    }
}
```
Note: output here capped at 5 rows for display purposes
::: 

Add up the numbers from the first column, without using the sum() function

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
counter <- 0
for (i in 1:nrow(data)) {
    counter <- counter + data[i,1]
}
print(counter)
```
::: 

<br> 

### if/else

Now we know how to loop over our data, next let's try filtering out some of the data based on certain conditions using `if/else` statements.

Let's start with printing rows where the blood pressure is above 80

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=F}
for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        print(data[i,])
    }
}
```
```{r, eval=T, echo=F}
for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        print(data[i,1:3])
    }
    if (i == 22) {
        break
    }
}
```
Note: output here capped at 5 rows and 3 columns for display purposes
::: 

Now let's investigate the BMI of the persons with blood pressure above 80. Print the BMI of those persons.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=F}
for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        print(data[i,6])
    }
}
```
```{r, eval=T, echo=F}
for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        print(data[i,6])
    }
    if (i == 22) {
        break
    }
}
```
Note: output here capped at 5 rows for display purposes.
::: 

Next, count how many people have blood pressure above 80?

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
high.bp <- 0
for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        high.bp <- high.bp + 1
    }
}
print(high.bp)
```
:::

[Tricky one!] Finally, do a comparison of the average BMI between people with blood pressure above 80, and people with blood pressure 80 and under.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
#| results: hold
above.bmi <- 0
above.count <- 0

below.bmi <- 0
below.count <- 0

for (i in 1:nrow(data)) {
    if (data[i,3] > 80) {
        above.bmi <- above.bmi + data[i,6]
        above.count <- above.count + 1
    } else {
        below.bmi <- below.bmi + data[i,6]
        below.count <- below.count + 1
    }
}
print(paste("Average BMI of BP > 80:", round(above.bmi/above.count, 2)))
print(paste("Average BMI of BP <= 80:", round(below.bmi/below.count, 2)))

```
:::

### Functions

Apart from using R's built in functions, you can also write your own functions! Writing your own functions is not a main part of this course, but you should know how they look like, and be able to modify them. The basic syntax for a functions is: 

```{r, eval=F}
function (<input parameters>) {
    <action>
    return <return value>
}
```

Start with writing a function called `add_two_values()`, that takes 2 values as input, sums them up, and returns the result.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
add_two_values <- function (x, y) {
    result <- x + y
    return (result)
}

add_two_values(4, 8)
```
:::


Create a function called `my_mean()` that takes a list as input, and calculates the mean of the list. You should not use the `mean()` function, but do it with a loop.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
my_mean <- function (vec) {
    sum <- 0
    count <- 0
    for (i in vec) {
        sum <- sum + i
        count <- count + 1
    }
    return (sum/count)
}
my_mean(c(1,2,3,4,5))
```
:::

Now use this function to calculate the mean of each column in your diabetes dataset.

::: {.callout-note collapse="true" appearance="minimal"}
## ![](img/play.svg){width=10}  Code
```{r, eval=T}
for (col in data) {
    print(my_mean(col))
}

```
:::

## Extra exercise material

A second pass at the same material, for extra practice. Try to solve each question on your own before peeking at the hints or solutions.

### Loops (extra)

- **QE1.1** Write a `for` loop that computes the factorial of 8 (that is, 1 * 2 * 3 * ... * 8) without using the built-in `factorial()` function. Store the result in a variable called `fact` and print it.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Initialise an accumulator *outside* the loop. For multiplication, the natural starting value is 1 (not 0). Loop over `1:8` and multiply the accumulator by each value.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
fact <- 1
for (i in 1:8) {
    fact <- fact * i
}
print(fact)
```
:::

- **QE1.2** Build the first 10 numbers of the Fibonacci sequence (1, 1, 2, 3, 5, 8, ...) using a `for` loop, and store them in a vector called `fib`. Print the resulting vector.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Start by creating a numeric vector of length 10 (for example with `numeric(10)`) and set the first two entries to 1. Then loop from `i = 3` to `10` and set `fib[i] <- fib[i-1] + fib[i-2]`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
fib <- numeric(10)
fib[1] <- 1
fib[2] <- 1
for (i in 3:10) {
    fib[i] <- fib[i-1] + fib[i-2]
}
print(fib)
```
:::

- **QE1.3** Compute the sum of the squares of the numbers from 1 to 20 (i.e. 1^2 + 2^2 + ... + 20^2) using a `for` loop. Do not use `sum()`. Print the result.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Initialise a counter at 0 outside the loop, then loop over `1:20` and add `i^2` to the counter on each iteration.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
total <- 0
for (i in 1:20) {
    total <- total + i^2
}
print(total)
```
:::

### if/else (extra)

- **QE2.1** Given a BMI value, classify it into one of four categories using an `if`/`else if`/`else` chain: `"underweight"` (BMI < 18.5), `"normal"` (18.5 <= BMI < 25), `"overweight"` (25 <= BMI < 30), and `"obese"` (BMI >= 30). Try it on `bmi <- 27.4`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Use the pattern:
```
if (cond1) {
    ...
} else if (cond2) {
    ...
} else if (cond3) {
    ...
} else {
    ...
}
```
Check the smallest range first, then build up.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
bmi <- 27.4
if (bmi < 18.5) {
    category <- "underweight"
} else if (bmi < 25) {
    category <- "normal"
} else if (bmi < 30) {
    category <- "overweight"
} else {
    category <- "obese"
}
print(category)
```
:::

- **QE2.2** Loop over the numbers from 1 to 20 and print, for each number: `"Fizz"` if it is divisible by 3, `"Buzz"` if it is divisible by 5, `"FizzBuzz"` if it is divisible by both, and otherwise the number itself.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
The modulo operator in R is `%%`. So `i %% 3 == 0` is `TRUE` when `i` is divisible by 3. Check the *both* case (divisible by 15) first, otherwise it will never be reached.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
for (i in 1:20) {
    if (i %% 15 == 0) {
        print("FizzBuzz")
    } else if (i %% 3 == 0) {
        print("Fizz")
    } else if (i %% 5 == 0) {
        print("Buzz")
    } else {
        print(i)
    }
}
```
:::

### Functions (extra)

- **QE3.1** Write a function called `summarise_vec()` that takes a numeric vector `vec` and returns a *list* with three named elements: `mean`, `min`, and `max`. Test it on `c(3, 7, 2, 9, 5)`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
You can construct a named list with `list(mean = ..., min = ..., max = ...)`. You may use the built-in `mean()`, `min()`, and `max()` functions here.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
summarise_vec <- function (vec) {
    return (list(
        mean = mean(vec),
        min  = min(vec),
        max  = max(vec)
    ))
}
summarise_vec(c(3, 7, 2, 9, 5))
```
:::

- **QE3.2** Write a function called `power()` that takes two arguments, `x` and `n`, and returns `x` raised to the power of `n`. Give `n` a *default value* of 2 so that calling `power(5)` returns 25, while `power(5, 3)` returns 125.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Default arguments are set inside the function definition, like `function (x, n = 2) { ... }`. Inside the body you can use `x^n`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
power <- function (x, n = 2) {
    return (x^n)
}
print(power(5))
print(power(5, 3))
```
:::

### Bioinformatics flavour (extra)

- **QE4.1** Write a function called `reverse_complement()` that takes a DNA string (made of the characters `A`, `C`, `G`, `T`) and returns its reverse complement. For example, `reverse_complement("ATGC")` should return `"GCAT"`. Test it on `"ATGCCGTA"`.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Split the string into individual characters with `strsplit(dna, "")[[1]]`. Loop over the characters and build the complement using `if`/`else if` (A <-> T, C <-> G). Finally, reverse the result with `rev()` and glue it back together with `paste(..., collapse = "")`.
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint (more)
Skeleton:
```
reverse_complement <- function (dna) {
    bases <- strsplit(dna, "")[[1]]
    comp  <- character(length(bases))
    for (i in seq_along(bases)) {
        if (bases[i] == "A") comp[i] <- "T"
        else if (bases[i] == "T") comp[i] <- "A"
        ...
    }
    return (paste(rev(comp), collapse = ""))
}
```
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
reverse_complement <- function (dna) {
    bases <- strsplit(dna, "")[[1]]
    comp  <- character(length(bases))
    for (i in seq_along(bases)) {
        if (bases[i] == "A") {
            comp[i] <- "T"
        } else if (bases[i] == "T") {
            comp[i] <- "A"
        } else if (bases[i] == "C") {
            comp[i] <- "G"
        } else if (bases[i] == "G") {
            comp[i] <- "C"
        } else {
            comp[i] <- "N"
        }
    }
    return (paste(rev(comp), collapse = ""))
}
reverse_complement("ATGCCGTA")
```
:::