---
title: "4.1 Basic statistics"
format:
  html:
    theme: yeti
---

If you want to describe a dice, probability is what you use. If you have two dice, still so… but as you gather enough dice and study their collective behavior, you effectively enter the realm of statistics — the field of "many stochastic variables". We also want to study data that does not stem from throwing dice or flipping coins. Here we will introduce you to the language used.

You have already studied statistics, but some sections here are likely still new to you. In particular, learn how to do statistical analysis using R.

## Review: Statistical distributions in general

**Introduction:** Watch this video on how we move from probability to statistics. *If you think this topic is rather basic, you can skip to doing the exercises below using R.*

- Video: *4.3 From probability to statistics* (Canvas/Kaltura)
- Watch this video: <https://www.youtube.com/watch?v=CfZa1daLjwo&list=PLaFfQroTgZnzbfK-Rie19FdV6diehETQy>
- Also watch a video showing in more detail how distributions behave (Canvas/Kaltura: *Demo of statistical distributions v2*).
- The demo used is browsable at <http://demo-statdist.serve.scilifelab.se>
- Optional: [this video](https://www.youtube.com/watch?v=b9a27XN_6tg&list=PLaFfQroTgZnzbfK-Rie19FdV6diehETQy&index=2) shows further examples of statistical distributions.
- Some possibly great solved examples: <https://www.youtube.com/watch?v=zJ8e_wAWUzE>

**Practice statistical distribution analysis using the demo (optional):**

- **Q1.1**. You have a stochastic variable X following a normal distribution, with mean 3 and standard deviation 5. What is the probability of X being greater than 7?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Standardize: how many standard deviations above the mean is 7? Then read off the upper tail of a standard normal.
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
z = (7 - 3) / 5 = 0.8. You want P(Z > 0.8) for Z ~ N(0,1).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
1 - pnorm(7, mean = 3, sd = 5)
```
Roughly 0.21.
:::

- **Q1.2**. Assume Z ~ N(0,2). What is P(Z > 10)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Watch the parameterisation: in R's `pnorm`, the second parameter of N(μ, σ) is the standard deviation, not the variance. Many textbooks use variance — check which is meant here.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
Reading N(0,2) as μ=0, σ=2:
```{r, eval=TRUE}
1 - pnorm(10, mean = 0, sd = 2)
```
Essentially zero (5 sigma away).
:::

- **Q1.3**. Assume Z ~ N(0,3). What is P(Z > 2 & Z < 5)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
P(a < Z < b) = F(b) - F(a), where F is the cumulative distribution function.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
pnorm(5, mean = 0, sd = 3) - pnorm(2, mean = 0, sd = 3)
```
:::

**Practice statistical distribution analysis using R (this is likely new to you!):**

- **Q2.1**. Look at the [R help for the normal distribution](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/Normal).

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
In R, `?Normal` or `?pnorm` opens the help page. Note the four functions: `dnorm` (density), `pnorm` (CDF), `qnorm` (quantile), `rnorm` (random).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
?Normal
?pnorm
```
:::

- **Q2.2**. Produce 3 random numbers that are normally distributed with mean 3 and sigma 5.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
The `r` prefix in `rnorm` stands for "random".
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
rnorm(3, mean = 3, sd = 5)
```
:::

- **Q2.3**. Assume X ~ N(0,1). What is P(X > 0)? (you don't need R to compute this; but check that the answer from R makes sense)

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
By symmetry the answer should be exactly 0.5. Use `pnorm` to confirm.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
1 - pnorm(0, mean = 0, sd = 1)
```
:::

- **Q2.4**. Assume X ~ N(0,2). What is P(X > 10)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Same as Q1.2 — `1 - pnorm(...)` for the upper tail.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
1 - pnorm(10, mean = 0, sd = 2)
```
:::

- **Q2.5**. Assume X ~ N(0,5). What is P(X > 10 & X < 20)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Subtract two CDF evaluations: P(10 < X < 20) = F(20) - F(10).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
pnorm(20, mean = 0, sd = 5) - pnorm(10, mean = 0, sd = 5)
```
:::

- **Q2.6.** Optional: verify the answers above using the visual demo.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Open <http://demo-statdist.serve.scilifelab.se>, pick a normal distribution, set μ and σ to match the question, then shade the region of interest.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
Visual check only — the shaded area in the demo should agree (within rounding) with the values returned by `pnorm`.
:::

- **Q2.7**. Assume X ~ Binom(theta=0.5, n=10); what is P(X=10)? Note that since the binomial distribution is discrete, you can here use the probability density function directly rather than using the cumulative density function (but either works).

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
For a discrete distribution, the "density" `dbinom(k, size, prob)` gives P(X = k) directly.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
dbinom(10, size = 10, prob = 0.5)
```
:::

- **Q2.8**. Verify above using basic probability theory. This case is a simple special case!

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
P(all 10 flips heads) = (1/2)^10.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
0.5^10
```
1/1024 ≈ 0.000977, matching `dbinom`.
:::

## The Poisson distribution

**Introduction:** This is one of the most fundamental distributions for handling discrete counts, as we get from DNA sequencers, or from Geiger counters for radiation. It arises whenever you have a memory-less process, emitting N events/second. A common example is traffic flow although in reality it follows much more complicated patterns.

- Video of the definition and some solved examples: <https://www.youtube.com/watch?v=jmqZG6roVqU>
- Video of some examples when a process is not Poisson. In general, whenever events are correlated for some reason, this property breaks down. This turns out especially important for single-cell analysis: <https://www.youtube.com/watch?v=sv_KXSiorFk>

**Practice the Poisson distribution:**

Assume X ~ Poisson(1).

- **Q3.1.** What is P(x=0)? P(x=1)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`dpois(k, lambda)` gives the Poisson probability mass at k.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
dpois(0, lambda = 1)
dpois(1, lambda = 1)
```
Both equal 1/e ≈ 0.368.
:::

- **Q3.2.** Plot a barplot of P(X=n) for n=0…10.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Build a vector with `dpois(0:10, lambda = 1)` and feed it to `barplot()`. Use `names.arg = 0:10` to label the bars.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
n <- 0:10
barplot(dpois(n, lambda = 1), names.arg = n,
        xlab = "n", ylab = "P(X = n)")
```
:::

- **Q3.3.** What is P(x ≥ 4)?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
You can either integrate, or you can relate this to P(x=0), P(x=1), P(x=2) and P(x=3).
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`ppois(q, lambda)` returns P(X ≤ q). So P(X ≥ 4) = 1 - P(X ≤ 3).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
1 - ppois(3, lambda = 1)
# equivalently:
ppois(3, lambda = 1, lower.tail = FALSE)
# or by summation:
1 - sum(dpois(0:3, lambda = 1))
```
:::

Counting DNA or RNA molecules is approximately Poisson-distributed. However, since we PCR-amplify them and produce copies (correlated!), this approximation breaks down if we pick too many molecules. Go to <https://demo-poisson.serve.scilifelab.se> and the tab "repeated sampling". You can set how many molecules you have originally in a tube, and how many molecules you sample using a sequencer.

- **Q3.4.** For 100 molecules, how many samples can you take while keeping doublets < 10%? Does this fit with your intuition?
    - number of points = how many molecules you can sample from
    - number of samples = how many molecules you pick
    - The table below shows the distribution of how many times a molecule has been sampled. If you don't sample at all, 100% of points will have numsamples=0.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Random seed controls the random sampling. Try different seeds to see how trying again affects the variation in your answer. Singlets are those with numsamples=1.
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Analytically: if you sample s molecules from N=100 with replacement, the expected occupancy per molecule is λ = s/N. The fraction of molecules hit ≥ 2 times is roughly 1 - (1+λ)e^(-λ). Solve for when this is 0.1.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=TRUE}
# Fraction of molecules sampled >= 2 times, as a function of total samples s,
# for N = 100 molecules.
N <- 100
s <- 1:100
lambda <- s / N
doublet_frac <- 1 - ppois(1, lambda)   # P(X >= 2)
data.frame(s = s, doublet_frac = round(doublet_frac, 3))[s %in% c(10, 20, 30, 50, 70), ]
```
Doublet rate crosses 10% somewhere around s ≈ 50–55 samples; the demo should agree.
:::

Optional related reading: [The birthday problem/paradox](https://en.wikipedia.org/wiki/Birthday_problem).

## Statistical testing in R

**Introduction:** Watch this video to learn how to perform basic statistical testing (Canvas/Kaltura: *4.1 Statistical_testing*).

Pay particular attention to:

- What is the mean and variance of the average?
- Why do we assume the average to have a normal distribution?
- How do you integrate the normal distribution to get p-values?

**Practice statistical testing using R:**

- Look up the [R help for t-testing](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test).
- **Q4.1.** Investigate [`diabetes.csv`](data/diabetes.csv.zip). Perform a t-test to check if the BMI of those with `Outcome=1` (i.e. those getting diabetes) is higher than those who do not. Is this a paired test?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Read the CSV with `read.csv()`, then split BMI by `Outcome` and feed the two vectors to `t.test()`. Use `alternative = "greater"` for a one-sided test.
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
"Paired" means each observation in group A has a natural partner in group B (e.g. before/after on the same patient). Here the two groups are different patients — so no.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
d <- read.csv("diabetes.csv")
bmi_pos <- d$BMI[d$Outcome == 1]
bmi_neg <- d$BMI[d$Outcome == 0]
t.test(bmi_pos, bmi_neg, alternative = "greater")
# equivalently using the formula interface:
t.test(BMI ~ Outcome, data = d, alternative = "less")  # 0 < 1 ordering
```
Not a paired test — the two groups are different individuals, with different sample sizes.
:::

- **Q4.2.** For "large" number of samples, you can also perform a non-parametric test, which makes fewer assumptions about the distribution of the samples. One such is `wilcox.test()`; how does it seem to compare to the t-test here?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Same call signature as `t.test()`. Compare the p-values; for reasonably symmetric, large-N data they should be in the same ballpark.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
d <- read.csv("diabetes.csv")
bmi_pos <- d$BMI[d$Outcome == 1]
bmi_neg <- d$BMI[d$Outcome == 0]
wilcox.test(bmi_pos, bmi_neg, alternative = "greater")
```
Both tests give very small p-values here — BMI is clearly higher in the diabetic group. The Wilcoxon makes no normality assumption but tends to be slightly less powerful when normality actually holds.
:::

- **Q4.3.** Optional: Perform similar tests on other variables in `diabetes.csv`; does it match your expectations?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Loop over the numeric columns and apply `t.test(col ~ Outcome, data = d)`. Watch out for columns where a value of 0 actually means "missing" (e.g. `Insulin`, `SkinThickness`).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
d <- read.csv("diabetes.csv")
vars <- setdiff(colnames(d), "Outcome")
pvals <- sapply(vars, function(v) {
  t.test(d[[v]] ~ d$Outcome)$p.value
})
sort(pvals)
```
Glucose, BMI, Age and Pregnancies all separate the two groups strongly — consistent with known risk factors for type 2 diabetes. Variables like `BloodPressure` and `SkinThickness` separate less clearly, partly because zeros in the raw data encode missing values.
:::
