---
title: "4.3 Linear models for statistics"
format:
  html:
    theme: yeti
---

- This module starts by recapping linear least squares from your intro statistics course. Learn how to solve such equations in R, as this is effectively what RNA-seq analysis boils down to.
- Much of the data in bioinformatics originate from sequencing and benefit from modelling using Poisson models. Linear models can be built upon Poisson models too but are difficult to solve using regular software. You will thus get to try RSTAN for whenever you need to perform advanced statistical modelling.

## Linear least squares

**Introduction:** If you have 4 data points and want to fit a line through them, you cannot do it using linear equation systems alone — it assumes that all 4 points all are neatly on the line, but the random noise added during measurement prevents it from being the case. Instead, we need to combine the linear model with the power of statistics to find the best possible solution to the equations. See this video to get a visual idea about why you want to minimize the "sum of squared errors":

<https://www.youtube.com/watch?v=eYTumjgE2IY>

The only thing the video fails to cover well is a bit of theory:

In the most basic case, we wish to solve for $k$ and $m$ in $y_i = k \cdot x_i + m$. If we have more than 2 data points, then we cannot find a solution in the basic linear algebra we have looked at.

Instead, we aim to fit a probabilistic equation, $y_i \sim \mathrm{Normal}(\mu = k \cdot x_i + m, \sigma)$ for some unknown $\sigma$. Because we aren't "sure" where points are as compared to the line, this system of equations can now be solved despite that you have more points than unknowns.

It turns out that this can be rewritten as the equivalent $y_i = k \cdot x_i + m + \epsilon$ where $\epsilon \sim \mathrm{Normal}(\mu = 0, \sigma)$. This is a more common formulation, and the one you see in this video.

However, the normal distribution is commonly ignored, as solving this probabilistic equation turns out equivalent to minimizing the squared error, or: $\sum_i \epsilon^2$ ; because engineers seldom need to handle anything but normal distribution, this fact is commonly glossed over. However, for now, you need not worry much about this, and focus instead on how to solve basic least squares linear models. We save the more general approach to later in this section!

- Read through this tutorial to see [how the `lm()` command is used to fit a linear model in R](https://www.datacamp.com/tutorial/linear-regression-R).
- To statistically test the coefficients and compute p-values, you will perform ANOVA-testing on the fitted model. This will tell if they are significantly different from 0 (i.e. they have a significant effect). We will not bother with the underlying math. Using ANOVA is simple enough that you can figure it out on your own. Otherwise, [here is an example](https://bookdown.org/steve_midway/DAR/understanding-anova-in-r.html).
- Optional: [this longer video may also help](https://www.youtube.com/watch?v=nk2CQITm_eo), but it doesn't really show any theory.
- Optional: [this great video covers the assumptions used in the model](https://www.youtube.com/watch?v=0MFpOQRY0rw) in much greater level of detail. If the assumptions do not hold up, you might end up with a really bad fit or wrong answer! It's fairly high level and easy to watch, but we are only covering one of the possible issues in this course.
- The interpretation of linear models is as before; but now you should also know about the statistical distribution of the fitted coefficients.
- Wikipedia on [linear least squares](https://en.wikipedia.org/wiki/Linear_least_squares).

**Practice solving linear least squares in R:**

Look at the R help [for `lm()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm) and [for `anova()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/anova).

Work on the provided dataset [`diabetes.csv`](data/diabetes.csv.zip).

- **Q1.1**. Using the function `lm()`, fit the model `BloodPressure ~ Age`.
    - Does age predict blood pressure?
    - Is it statistically significant?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Load the data with `read.csv()`, then pass the formula to `lm()`. Inspect the fit using `summary()` or `anova()` to read off the coefficient and its p-value.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
d <- read.csv("diabetes.csv")
fit <- lm(BloodPressure ~ Age, data = d)
summary(fit)
anova(fit)
```
Age has a positive coefficient — older subjects tend to have higher blood pressure — and the effect is statistically significant (small p-value).
:::

- **Q1.2**. Fit the model `BloodPressure ~ Age + Glucose`. Does this appear to affect the situation?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Add `+ Glucose` to the right-hand side of the formula. Compare the p-value and coefficient of `Age` with what you got in Q1.1.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit2 <- lm(BloodPressure ~ Age + Glucose, data = d)
summary(fit2)
anova(fit2)
```
Adding Glucose all but erases the Age effect: the Age coefficient drops from about 0.22 to 0.08 and is no longer statistically significant (p ≈ 0.17), while Glucose itself is highly significant. Because Glucose is correlated with Age, much of what looked like an "Age effect" in Q1.1 is actually reassigned to Glucose once both are in the model. A single correlated predictor was enough to make Age look irrelevant.
:::

- **Q1.3**. Fit a model using all available variables. This can be done with the shorthand: `BloodPressure ~ .`
    - Is age still as significant? What do you think happened here?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
The dot `.` on the right-hand side means "every other column in `data`". Refit and look again at the row for `Age` in the summary table.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit_all <- lm(BloodPressure ~ ., data = d)
summary(fit_all)
anova(fit_all)
```
Now something striking happens: the Age coefficient does not just lose significance, it **flips sign** — it becomes about −0.59 and is again highly significant (p ≈ 0). So depending on what else is in the model, "more age" appears to *raise* blood pressure (Q1.1) or *lower* it (here). `Pregnancies`, `Glucose`, `Insulin`, `BMI` and `Outcome` come out significant, while `SkinThickness` and `DiabetesPedigreeFunction` do not. The reversal happens because `Pregnancies` is strongly correlated with `Age` (≈ 0.78): in the simple model Age was largely standing in for Pregnancies, and once Pregnancies (and the other correlated drivers) are in the model the *direct* Age effect points the other way. A coefficient — even its sign — can change completely once you add correlated covariates; an apparent association in a simple model does not by itself establish causation.
:::

- **Q1.4** Optional: Plot Age vs some key variables to help your intuition about what is going on.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
A quick `plot(d$Age, d$Glucose)` is enough — repeat for Pregnancies and a few other columns. `pairs(d)` gives all pairwise scatterplots at once.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
plot(d$Age, d$Glucose)
plot(d$Age, d$Pregnancies)
pairs(d)
```
Older subjects tend to have many more Pregnancies and somewhat higher Glucose (cor(Age, Pregnancies) ≈ 0.78, cor(Age, Glucose) ≈ 0.34). Note that not every variable tracks Age — BMI, for instance, is essentially uncorrelated with Age here (cor ≈ 0.09). Because Pregnancies in particular moves so tightly with Age, the simple `BloodPressure ~ Age` model is largely picking up the Pregnancies (and Glucose) effect rather than a direct effect of age itself — which is exactly why the Age coefficient collapsed and then reversed as those variables were added.
:::

We just watched this happen: Age looked like a clear positive predictor on its own, lost all significance once Glucose was added, and then turned significantly *negative* once the strongly correlated Pregnancies variable entered the model. Adding a correlated predictor can shrink, inflate, or even reverse a coefficient — the most dramatic version of this is [Simpson's paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox). The more variables you throw at a prediction, the more careful you must be about causality. Correlations alone cannot be used to infer causation, especially not in biology!

## Handle other statistical distributions using (R)STAN

**Introduction:** The normal distribution turns out to have nice mathematical properties, which is a reason it is commonly used. However, sequencing data is discrete, i.e., you have N molecules, and N is never a negative number. Thus bioinformaticians these days frequently need to use some type of Poisson distribution. We will get back to this topic in more detail in the next module.

See this video on how linear models can be more generally formulated, and how we can use STAN and R to solve them numerically (Canvas/Kaltura: *4.4 Linear models with STAN*).

- Optional: Various STAN examples [here](https://hedibert.org/wp-content/uploads/2021/02/stan-rstan-examples.html).

Note that STAN primarily is meant for Bayesian statistics (next module!) but it is very capable of computing maximum likelihood estimates for frequentist statistics (i.e. what we have done until now).

**Practice using STAN:**

- **Q2.1**. In the file [`frog_stan.zip`](data/frog_stan.zip), check out **frog.R**, **frog.stan**, and make this example run on your computer. If you struggle to open the stan-file in RStudio, you will likely find it easiest with the "File → open file…" command in the RStudio menus.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Make sure the `rstan` package is installed (`install.packages("rstan")`). Unzip the archive into a directory, set that directory as the working directory in RStudio, then open **frog.R** and source it.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
# 1. unzip frog_stan.zip
# 2. setwd() to the unzipped folder, or open the project there
# 3. install.packages("rstan") if needed
# 4. open frog.R in RStudio and run it line by line
library(rstan)
source("frog.R")
```
The script reads the frog counts, compiles **frog.stan**, runs the sampler, and prints / plots the result.
:::

- **Q2.2**. Study the code and make yourself familiar with what it does. Note that **frog.stan** contains code that you should have a look at as well.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Look at the `data`, `parameters`, and `model` blocks in **frog.stan**. What are the observations? What unknowns is STAN solving for? How is the rate at intermediate times computed from the two endpoint rates?
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
The observed frog counts are modelled as Poisson-distributed: `y[i] ~ Poisson(lambda[i])`. The rate `lambda[i]` is not free for each time point — instead it is linearly interpolated between two endpoint rates `lambda_start` and `lambda_end` using `lambda[i] = (1-time[i])*lambda_start + time[i]*lambda_end`, where `time` runs from 0 to 1 across the observation times. At `time = 0` this gives `lambda_start` and at `time = 1` it gives `lambda_end`. STAN samples (or maximises the likelihood over) `lambda_start` and `lambda_end` given the data.
:::

**Note on the linear model / interpolation used in the model:** This code has x-values in the range 0…1. This simplifies the interpolation code. Consider $f(x) = x a + (1 - x) b$. Note that $f(0) = b$ and $f(1) = a$. Thus these define the values at either extreme. A more general way of doing linear interpolation is however $f(x) = \frac{x}{N} a + \left(1 - \frac{x}{N}\right) b$. With this definition, $f(0) = b$ and $f(N) = a$. Thus it would be suitable for X being 0…N instead.

- **Q2.3**: Plot the number of frogs over time. What does the trend look like?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
A plain `plot(frogdat$time, frogdat$y)` will do; add `type = "b"` to connect the points if you like.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
frogdat <- read.csv("frogs.csv")
plot(frogdat$time, frogdat$y, type = "b", xlab = "time", ylab = "frog count")
```
The counts trend upward over time — the frog population looks like it is increasing.
:::

- **Q2.4**: After running the provided code, does the model fit the data?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Compare the posterior (or MLE) values of `lambda_start` and `lambda_end` against the data, and overlay the fitted line `lambda = (1-time)*lambda_start + time*lambda_end` on the scatterplot from Q2.3.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
print(fit)
# overlay the interpolated rate on the observed counts
```
Yes — the linear interpolation between the two endpoint rates captures the increasing trend well (the fitted rate rises from `lambda_start` ≈ 19.4 to `lambda_end` ≈ 24.5), and the observed counts scatter around the fitted curve as expected for Poisson noise.
:::

- **Q2.5: Optional.** In the STAN-code, you see that there is a limit on lambda to be ≥ 0. Why is this? (without limits, lambda is initially taken from a normal distribution around 0)

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
What does `lambda` mean in a Poisson distribution? Is there any value it is not allowed to take?
:::

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Think about what happens to `Poisson(lambda)` when `lambda` is negative — the density (and its log) is not defined.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
`lambda` is the rate parameter of a Poisson distribution and must be non-negative — a Poisson with negative rate is undefined. Without the `<lower=0>` constraint, STAN initialises parameters from a normal distribution centred on 0 and would happily propose negative `lambda` values; the likelihood would then involve `log(lambda)` of a non-positive number, producing `NaN` or `-Inf`, and the sampler fails to start. The constraint reparameterises so that only valid values are ever proposed.
:::

## Extra exercise material

A second pass at the same material, for extra practice. Same conventions as above — try each question first, then peek.

### Linear models on diabetes.csv (extra)

Work again on [`diabetes.csv`](data/diabetes.csv.zip).

- **QE1.1**. Fit `Glucose ~ BMI + Age`. Which predictor has the larger effect per unit, and which is more statistically significant? Are the two questions the same?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Read both the `Estimate` column and the `Pr(>|t|)` column of `summary(fit)`. A big coefficient is not the same as a small p-value — the standard error matters too. Also remember `BMI` and `Age` are on very different scales.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
d <- read.csv("diabetes.csv")
fit <- lm(Glucose ~ BMI + Age, data = d)
summary(fit)
```
The size of an estimate is measured in units of the predictor (mmol/kg/m² vs. mmol/year), so the raw numbers cannot be compared directly. Significance depends on `Estimate / Std.Error`, not on the magnitude of `Estimate` alone. To compare effect sizes fairly you would have to standardise the predictors first (`scale()`).
:::

- **QE1.2**. Fit `Insulin ~ Age + BMI`. Then refit it as `Insulin ~ Age * BMI` (note the `*`). What new term appears, and what does it mean?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`Age * BMI` expands to `Age + BMI + Age:BMI`. The `Age:BMI` row is the interaction.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit_add <- lm(Insulin ~ Age + BMI, data = d)
fit_int <- lm(Insulin ~ Age * BMI, data = d)
summary(fit_int)
```
The extra `Age:BMI` term asks whether the effect of BMI on insulin depends on Age (equivalently, whether the effect of Age depends on BMI). A non-zero interaction means the two predictors do not simply add.
:::

- **QE1.3**. The column `Outcome` is binary (0 / 1, diabetic or not). Fit `Outcome ~ Glucose + BMI + Age` first with `lm()`, then with `glm(..., family = binomial)`. Why is `lm()` the wrong tool here?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
A linear model assumes the response is continuous and the residuals are normal. With a 0/1 outcome the fitted values can fall outside `[0, 1]` and the residuals are clearly not normal. Logistic regression (`glm` with `family = binomial`) models the log-odds instead.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit_lm  <- lm(Outcome ~ Glucose + BMI + Age, data = d)
fit_glm <- glm(Outcome ~ Glucose + BMI + Age, data = d, family = binomial)
summary(fit_glm)
range(predict(fit_lm))                       # can go <0 or >1
range(predict(fit_glm, type = "response"))   # bounded in [0, 1]
```
`lm()` will produce predicted "probabilities" outside `[0, 1]` and homoscedasticity is violated. `glm(..., family = binomial)` fits a logistic regression — coefficients are now on the log-odds scale, but the predicted probabilities are bounded and the likelihood is appropriate for binary data.
:::

- **QE1.4**. Fit `Outcome ~ .` with `glm(family = binomial)`. Which variables come out as the strongest predictors of diabetes? Compare with what `lm(Outcome ~ .)` would have told you.

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
Use `summary(fit)` and rank by p-value (or `|z value|`). The signs of the coefficients tell you the direction of association on the log-odds scale.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit_full <- glm(Outcome ~ ., data = d, family = binomial)
summary(fit_full)
```
Typically `Glucose`, `BMI`, and `Pregnancies` come out as strongly associated with `Outcome`. The relative ordering and significance from `glm()` is the trustworthy version; the `lm()` p-values are biased because the model assumptions are violated.
:::

### Diagnostics and assumptions (extra)

- **QE2.1**. Re-fit `BloodPressure ~ Age + BMI + Glucose` and run `plot(fit)`. What four diagnostic plots come up, and what is each one checking?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
By default `plot.lm` shows: residuals vs fitted, Normal Q-Q of residuals, scale-location (sqrt-abs-residuals vs fitted), and residuals vs leverage with Cook's distance.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit <- lm(BloodPressure ~ Age + BMI + Glucose, data = d)
par(mfrow = c(2, 2))
plot(fit)
```
- Residuals vs fitted: checks linearity and constant variance — looking for no trend.
- Normal Q-Q: checks normality of residuals — points should lie on the diagonal.
- Scale-location: checks homoscedasticity — flat horizontal line is good.
- Residuals vs leverage: flags individual points with high influence (Cook's distance).
:::

- **QE2.2**. From the Q-Q plot, do the residuals look normal? And if the scale-location plot fans out (residuals get bigger as fitted values grow), what is the technical name for that problem and what is one common fix?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
"Fanning out" = variance depends on the mean. Look up *heteroscedasticity*. Common fixes: transform the response (`log`, `sqrt`), use weighted least squares, or move to a GLM with a variance function that matches the data (Poisson, gamma, ...).
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
qqnorm(residuals(fit)); qqline(residuals(fit))
plot(fit, which = 3)   # scale-location
```
Fanning residuals indicate **heteroscedasticity** — the constant-variance assumption is violated. A log- or sqrt-transform of the response, weighted least squares, or a GLM with an appropriate family (e.g. Poisson for count data) typically helps. Non-normal residuals can also point to outliers or a wrong link function.
:::

### ANOVA vs. summary (extra)

- **QE3.2**. Compare two nested models with `anova(fit_small, fit_big)`. What null hypothesis is being tested?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
With two model arguments, `anova()` performs an F-test of whether the extra terms in the bigger model significantly improve the fit.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
fit_small <- lm(BloodPressure ~ Age, data = d)
fit_big   <- lm(BloodPressure ~ Age + BMI + Glucose, data = d)
anova(fit_small, fit_big)
```
H0: the extra coefficients (`BMI`, `Glucose`) are jointly zero — i.e. the smaller model fits just as well. A small p-value rejects H0 and favours the bigger model.
:::

### STAN (extra)

These exercises build on [`frog_stan.zip`](data/frog_stan.zip).

- **QE4.1**. Modify **frog.stan** so that instead of linearly interpolating between two endpoint rates, `log(lambda)` is linear in time: `log(lambda[i]) = a + b * t[i]`. What changes in the `parameters` block, and do you still need `<lower=0>`?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
On the log scale `lambda` is automatically positive, so the constraint is no longer required for `a` and `b`. Pass `t` as part of `data`, drop the interpolation arithmetic, use `poisson_log`.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
# data    { int N; vector[N] t; int y[N]; }
# parameters { real a; real b; }
# model   { y ~ poisson_log(a + b * t); }
```
Because `exp(a + b*t) > 0` for all real `a`, `b`, the `<lower=0>` constraint is unnecessary. `b < 0` would correspond to an exponentially declining frog population.
:::

- **QE4.2**. Run **frog.R** twice: once with the default MCMC sampler (`stan(...)`) and once with maximum-likelihood (`optimizing(...)`). How should you interpret the two outputs differently?

::: {.callout-tip collapse="true" appearance="minimal"}
## Hint
`stan()` gives you a **posterior distribution** over `a` and `b` — point estimates, credible intervals, full samples. `optimizing()` gives you a single point: the **maximum-likelihood (or MAP) estimate**, no uncertainty unless you bootstrap or use the Hessian.
:::

::: {.callout-note collapse="true" appearance="minimal"}
## Solution
```{r, eval=FALSE}
library(rstan)
sm  <- stan_model("frog.stan")
mle <- optimizing(sm, data = stan_data)   # point estimate
post <- sampling(sm, data = stan_data)    # full posterior
mle$par
print(post)                                # mean, sd, 2.5%/97.5% per parameter
```
The MLE answers "what single value of `(a, b)` makes the data most likely?". The Bayesian posterior answers "given the data and the prior, what is the full distribution of plausible `(a, b)` values?". For reporting uncertainty, intervals from the posterior are usually more honest; the MLE is faster and is what classical `lm()` / `glm()` give you implicitly.
:::
